Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-23 paper.tex

\subsection{Interoperability: vcztools}
To provide compatibility with existing workflows we have 
developed the ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality.
[TODO DETAILS]
% (such as 
% \texttt{view} and \texttt{query}) on a VCF Zarr dataset.
% Such a tool would speed up some common queries by orders 
% of magnitude, and reduce the need for user orchestration of 
% operations among manually split VCF chunks (large VCF datasets
% are typically split into hundreds of files; see the Genomics
% England case study). Datasets could then be hosted in cloud
% object stores, while still presenting file-like semantics
% for existing workflows. This could provide an evolutionary path,
% allowing established analysis workflows to coexist with new Zarr-native
% approaches, working from the same primary data.

>>>>

\subsection{Interoperability: vcztools}
To provide compatibility with existing workflows we have 
developed the ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality.
Vcztools is intended to be a drop-in replacement for \texttt{bcftools},
and compatibility being a primary objective.
Vcztools is primarily intended to provide an evolutionary path for 
VCF Zarr adoption,
allowing established analysis workflows to coexist with new Zarr-native
approaches, working from the same primary data.
Vcztools currently supports Zarr stored in the local file system
and Zip archives, and support for major cloud stores will be added in the near
future. [JK: note this is a pretty simple feature, maybe we should just try 
to add it?]
% Mention Pyparsing?
It is written in a mixture of Python and C, using
PyRanges~\citep{stovner2020pyranges} to perform efficient range queries.

The initial version of vcztools implements a subset of the \texttt{query}
and \texttt{view} commands. We provide an example based on the simulated 
dataset of $10^5$ samples of Fig~\ref{fig-data-storage} to illustrate 
the functionality and 
performance of vcztools. First, we extract the position column 
using the \texttt{query} command. Using bcftools
(\texttt{bcftools query -f '\%POS\textbackslash n' chr21\_10\_5.bcf})
it took 8m57s to extract the position values for the 2,365,367
variants from the BCF file. 
Using vcztools
(\texttt{vcztools query -f '\%POS\textbackslash n' chr21\_10\_5.zarr})
required 51s (note that this operation could be performed much
more quickly and may be worth optimising for, see 
Fig~\ref{fig-column-extract}). This illustrates the general principle
that vcztools only retrieves information necessary to fulfil a particular 
query, which can have a significant impact with large datasets.

To form a practical replacement for bcftools in production pipelines
it is important that vcztools can produce VCF text at a reasonable
rate. To do this we measure the time taken to generate VCF text for the 
last 10,000 variants in the dataset (using the previously extracted 
position column to determine the coordinates, \texttt{vcztools view
chr21\_10\_5.zarr -r1:47958263-}). This took 30s, at rate of 124.3 MiB
of VCF text produced per second. The bcftools command on the corresponding
BCF file took 25s (152.64 MiB/s) and on the VCF file it took 63s (60.57 MiB/s).
Thus, vcztools view can produce VCF at an intermediate rate between 
bcftools reading from BCF and VCF files.

Vcztools is primarily intended to used as a replacement for bcftools 
in existing workflows based on Unix pipes, where VCF text is sequentially
filtered and processed by different processes in the classical
Unix manner. In this context the most important property is that vcztools
can produce VCF text as quickly as other processes in the pipeline can consume
it. To measure the rate at which VCF can be parsed, we wrote a simple
program that reads a VCF file and parses it line-by-line using \texttt{htslib}
APIs~\citep{bonfield2021htslib}. For the 10,000 variant subset discussed
above (3.73 GiB) this took 31s, giving a parse rate of 123.1 MiB/s
(almost identical to the write-rate of vcztools above). Given this limiting
factor on overall throughput,
it is unlikely that \texttt{vcztools view} would be a bottleneck 
in practical processing pipelines.
